Bayesian Statistical Modeling with Predictors from LLMs

### 2024-06-13 04-paper/paper-LLM-probmod.tex
fixed typo in section heading

\usepackage{mypackages}
\bibliography{~/Desktop/icloud/MyRefGlobal}
\usepackage{mycommands}

@@

\begin{figure}[t]
  \centering
  \includegraphics[width = 0.95\textwidth]{00-pics/condition-item-level-prediction.png}
  \caption{
    Schematic representation of key conceptual differences between different types of predictive models.
    Standard statistical models, like hierarchical regression models, typically make predictions for aggregate data (e.g., at the condition level), and add random offsets for item-level variation.
    More sophisticated models, like some probabilistic cognitive models, may holistically combine information from an aggregate level (task, condition) with specific information about items.
    LLMs, in contrast, first and foremost give prediction about each individual item and must rely on proper aggregation to arrive at condition-level predictions.
  }
  \label{fig:stats-model-types}
\end{figure}

@@

The paper is structured as follows.
Section~\ref{experiment-reference-games} describes an experiment with human participants with a text-based reference game.
Section~\ref{sec:model-pred-from} introduces the Rational Speech Act (RSA) model for the human data from the reference game experiment.
Section~\ref{sec:item-level-pred} exposes a statistical model for item-level predictions derived from LLM scores and investigates whether this adequately captures the human data at the item-level based on scores from GPT-3.5 \citep{BrownMann2020:Language-Models}.
Section~\ref{llm-predictions-for-reference-games} discusses different ways of deriving probabilistic predictions from LLMs at the condition-level and compares them against the human data and each other.
Finally, Section~\ref{sec:gener-other-llm} explores whether previous results generalize to other LLM backends by investigating different versions of LLaMA2 \citep{TouvronLavril2023:LLaMA:-Open-and}.\footnote{
  All data and code is available at \url{https://osf.io/f6j3a/?view_only=5e820cc8bbee4549aed58dc252ba61b9}.
}

@@

  \includegraphics[width = 0.8\textwidth]{00-pics/reference-game.png}

@@

\begin{figure}[t]
  \centering
  \includegraphics[width = 0.9 \textwidth]{00-pics/RSA-example.png}
  \caption{
    Example of predictions from the RSA model.
    The semantic meaning is shown as a matrix of binary truth-values.
    The policies of literal listener, pragmatic speaker and listener are calculated for uniform priors over states (referents) for $\alpha=1$, and are shown as row-stochastic matrices.
  }
  \label{fig:RSA-example}
\end{figure}

@@

    \includegraphics[width=0.9\linewidth]{00-pics/GPT-PPC-alpha-eps-model.pdf}

@@

\begin{figure}[t]
  \centering
  \includegraphics[width=\textwidth]{00-pics/GPT-closeness-target-by-alpha-item-level.pdf}
  \caption{
    Predicted probability of highest-scoring answer category, averaged over items, for different values of softmax parameter $\alpha$ (black lines).
    The gray-shaded area indicates the posterior 95\% credible interval for $\alpha$, and the implied probabilistic prediction.
    For reference, the target probability under a random choice strategy and under the ``winner-takes-all'' (WTA) strategy are shown with dashed lines.
    For credible values of $\alpha$, the means of predicted probabilities for the target option are clearly distinct from the WTA strategy.
  }
  \label{fig:closeness-target-item-level}
\end{figure}

@@

  \includegraphics[width = 0.9\linewidth]{00-pics/GPT-item-combined-obs-pred.pdf}

@@

\begin{figure}
  \centering
  \includegraphics[width=0.9\textwidth]{00-pics/measures-overview.png}
  \caption{
    Schematic overview over three different methods of obtaining condition-level predictions by aggregating item-level information.
    Methods differ in the kind of item-level information they take into account, which entails differences in the order in which aggregation, transformation to probabilities and parameterized scaling occur.
  }
  \label{fig:measures-overview}
\end{figure}

@@

\begin{figure}[t]
  \centering
  \includegraphics[width=0.9\textwidth]{00-pics/vPPC_all_LLaMA_models.pdf}
  \caption{Visual posterior predictive checks on condition-level data for different models of the LLaMA family, paired with different aggregation strategies.}
  \label{fig:LLaMA-results-vPPC}
\end{figure}

@@

\begin{figure}[H]
  \centering
  \includegraphics[width = 0.8\textwidth]{00-pics/refgame-production.png}

@@

\begin{figure}[H]
  \centering
  \includegraphics[width = 0.8\textwidth]{00-pics/refgame-interpretation.png}

>>>>

\usepackage{mypackages}
\bibliography{MyRefGlobal}
\usepackage{mycommands}

@@

\begin{figure}[t]
  \centering
  \includegraphics[width = 0.95\textwidth]{pics-final/condition-item-level-prediction.png}
  \caption{
    Schematic representation of key conceptual differences between different types of predictive models.
    Standard statistical models, like hierarchical regression models, typically make predictions for aggregate data (e.g., at the condition level), and add random offsets for item-level variation.
    More sophisticated models, like some probabilistic cognitive models, may holistically combine information from an aggregate level (task, condition) with specific information about items.
    LLMs, in contrast, first and foremost give prediction about each individual item and must rely on proper aggregation to arrive at condition-level predictions.
  }
  \label{fig:stats-model-types}
\end{figure}

@@

The paper is structured as follows.
Section~\ref{experiment-reference-games} describes an experiment with human participants with a text-based reference game.
Section~\ref{sec:model-pred-from} introduces the Rational Speech Act (RSA) model for the human data from the reference game experiment.
Section~\ref{sec:item-level-pred} exposes a statistical model for item-level predictions derived from LLM scores and investigates whether this adequately captures the human data at the item-level based on scores from GPT-3.5 \citep{BrownMann2020:Language-Models}.
Section~\ref{llm-predictions-for-reference-games} discusses different ways of deriving probabilistic predictions from LLMs at the condition-level and compares them against the human data and each other.
Finally, Section~\ref{sec:gener-other-llm} explores whether previous results generalize to other LLM backends by investigating different versions of LLaMA2 \citep{TouvronLavril2023:LLaMA:-Open-and}.\footnote{
  All data and code is available at this OSF repository:  \href{https://osf.io/f6j3a/?view_only=5e820cc8bbee4549aed58dc252ba61b9}{https://osf.io/f6j3a/?view\_only=5e820cc8bbee4549aed58dc252ba61b9}.
}

@@

  \includegraphics[width = 0.8\textwidth]{pics-final/reference-game.png}

@@

\begin{figure}[t]
  \centering
  \includegraphics[width = 0.9 \textwidth]{pics-final/RSA-example.png}
  \caption{
    Example of predictions from the RSA model.
    The semantic meaning is shown as a matrix of binary truth-values.
    The policies of literal listener, pragmatic speaker and listener are calculated for uniform priors over states (referents) for $\alpha=1$, and are shown as row-stochastic matrices.
  }
  \label{fig:RSA-example}
\end{figure}

@@

    \includegraphics[width=0.9\linewidth]{pics-final/GPT-PPC-alpha-eps-model.pdf}

@@

\begin{figure}[t]
  \centering
  \includegraphics[width=\textwidth]{pics-final/GPT-closeness-target-by-alpha-item-level.pdf}
  \caption{
    Predicted probability of highest-scoring answer category, averaged over items, for different values of softmax parameter $\alpha$ (black lines).
    The gray-shaded area indicates the posterior 95\% credible interval for $\alpha$, and the implied probabilistic prediction.
    For reference, the target probability under a random choice strategy and under the ``winner-takes-all'' (WTA) strategy are shown with dashed lines.
    For credible values of $\alpha$, the means of predicted probabilities for the target option are clearly distinct from the WTA strategy.
  }
  \label{fig:closeness-target-item-level}
\end{figure}

@@

  \includegraphics[width = 0.9\linewidth]{pics-final/GPT-item-combined-obs-pred.pdf}

@@

\begin{figure}
  \centering
  \includegraphics[width=0.9\textwidth]{pics-final/measures-overview.png}
  \caption{
    Schematic overview over three different methods of obtaining condition-level predictions by aggregating item-level information.
    Methods differ in the kind of item-level information they take into account, which entails differences in the order in which aggregation, transformation to probabilities and parameterized scaling occur.
  }
  \label{fig:measures-overview}
\end{figure}

@@

\begin{figure}[t]
  \centering
  \includegraphics[width=0.9\textwidth]{pics-final/vPPC_all_LLaMA_models.pdf}
  \caption{Visual posterior predictive checks on condition-level data for different models of the LLaMA family, paired with different aggregation strategies.}
  \label{fig:LLaMA-results-vPPC}
\end{figure}

@@

\begin{figure}[H]
  \centering
  \includegraphics[width = 0.8\textwidth]{pics-final/refgame-production.png}

@@

\begin{figure}[H]
  \centering
  \includegraphics[width = 0.8\textwidth]{pics-final/refgame-interpretation.png}
